Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-02 linear_iso.tex
updating main part of paper

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant. We  
first examine isothermal perturbations ($\beta = 0$) before
considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we introduce a number
of approximations below that allows a simple governing equation to be
derived and compared with previous works.

\subsection{The linear problem in the vertically global
  shearing box} 
We connect the above linearized equations to the vertically global
shearing box (VGSB) formalism developed by \cite{mcnally14}. The VGSB
is a generalization of the standard shearing box \citep{goldreich65}
by allowing the shear flow in the shearing box to be height-dependent.

The linear problem in the VGSB framework can be derived from 
Eq. \ref{lin_mass}---\ref{lin_energy} by setting $\hat{g}_c=0$, since 
the VGSB, like the standard shearing box, is homogeneous in the radial 
direction in equilibrium. The resulting linear problem is very similar
to that for local axisymmetric waves in accretion disks already
discussed by \cite{lubow93}, but modified by the presence of vertical
shear and thermal relaxation. This is convenient as it allows us to
directly apply their solution methods.

@@

We further discuss caveats of the VGSB in Appendix \ref{global_corr}.
However, we find that for perturbations with small radial lengthscales
($k_xH_\mathrm{iso}\gg 1$) and fast thermal relaxation ($\beta\ll1$), which
characterizes the VSI, the VGSB framework is adequate for the purpose
of a simple analytical discussion.

% In effect, we are studying dynamics in the standard shearing box
% subject to an additional azimuthal force $-v_zr_0 d \Omega_0/dz$.

% $d\Omega_0/dz$ is a free parameter (as are $\Omega_0$ and $S_0$). 
% This is different to the global problem because in that case, the
% vertical shear is a result of the disk radial structure, which is
% ignored in the VGSB equilibria. 
% In effect, we are studying dynamics in the standard shearing box
% subject to an additional azimuthal force $-v_zr_0 d \Omega_0/dz$. This
% force is inspired by the vertical shear in the global disk, and its
% functional form is taken from global equilibrium.

% \subsection{Limitations}
% Before using the above equations to study the VSI, it is very important to be
% aware of several limitations of the VGSB model. The VGSB momentum  
% equations have the same form as that for the standard shearing box 
% \citep{goldreich65}, except for the appearence of the term 
% $v_zr_0d\Omega_0/dz$. This term appears in the global stability
% problem and is the only term associated with vertical shear that is 
% ported from the global disk into the local model.

% In Appendix \ref{global_corr} we make the connection between the
% linear problem in the VGSB to that in the global disk model. We
% emphasize that not all calculations in the local model can be 
% consistently interpreted in the global context. However, for
% perturbations with small radial lengthscales (large $|k_x|$) and fast
% thermal relaxation ($\beta\ll1$), which characterizes the VSI, the
% VGSB framework is adequate.

\subsection{Approximate governing equation}\label{approx_gov}
We now derive a governing equation with further approximations:
\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$.

>>>>

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. We  first examine isothermal perturbations ($\beta =
0$) before considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.

\subsection{Relation to the vertically global
  shearing box} 
We connect the above linearized equations to the vertically global
shearing box (VGSB) formalism developed by \cite{mcnally14}. The VGSB
is an extension of the standard shearing box \citep{goldreich65} to
allow the shear flow in the shearing box to be height-dependent.  
The difference between the radially-local approximation and the VGSB is
that the radial disk structure is ignored in the latter.  That is, the
equilibrium VGSB is homogeneous in the radial direction, as in the
standard shearing box.

Thus, the linear problem in the VGSB can be derived from
Eq. \ref{lin_mass}---\ref{lin_energy} by setting 
$\hat{g}_c=0$.  
The resulting linear problem is very similar
to that for local axisymmetric waves in accretion disks already
discussed by \cite{lubow93}, but modified by the presence of vertical
shear and thermal relaxation. This is convenient as it allows us to
directly apply their solution methods.

@@

We further discuss caveats of the VGSB in Appendix \ref{global_corr}.
However, we find that for perturbations with small radial lengthscales
($k_xH\gg 1$) and fast thermal relaxation ($\beta\ll1$), which
characterizes the VSI, the VGSB framework is adequate for the purpose
of a simple analytical discussion.

\subsection{Governing equation in the VGSB}\label{approx_gov}
We now derive a governing equation with further approximations:
\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$.

### 2015-03-02 paper.tex
updating main part of paper

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the
  disk  midplane, which is generally expected in 
  astrophysical disks. However, the VSI requires a short thermal
  timescale. A quantitative assessment of the thermodynamic
  requirement for the VSI is thus desirable for application to
  realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include an energy
  equation with a thermal relaxation term that restores the disk
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ away from zero
  dramatically reduces VSI growth rates because this generally
  introduces a strongly stabilizing vertical entropy gradient. Our
  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify `fast' 
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. This 
  suggests the VSI is dynamically important in the outer parts of
  protoplanetary disks.  
\end{abstract}

@@

%recent work on vsi 
One possibility is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. The VSI is the
disk analog of the Goldreich-Schubert-Fricke instability
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple explanation for this 
instability: an exchange of fluid parcels while conserving their
angular momenta individually, may result in a new configuration of
lower energy, if the exchange occurs over a sufficiently large
vertical displacement.

@@

The VSI has been studied analytically in local approximation
\citep{urpin98,urpin03}. Simulations of the  
VSI have been performed by \cite{arlt04}, and more recently by 
\cite{nelson13}. In these simulations the authors indeed observe the
growth of vertically-global but radially-localized perturbations when
the equilibrium rotation varied with height.

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions are stabilized by
buoyancy if the disk is sub-adiabatically stratified, as required for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P_\mathrm{orb}$, \cite{nelson13} find that
$t_c\lesssim 0.01P_\mathrm{orb}$ was needed to observe instability. That is, the
temperature profile of the disk essentially remains unchanged during
the disk evolution. This has been confirmed by 
\cite{stoll14}, who find the 
VSI is only sustained in radiation hydrodynamic simulations 
with the inclusion of heating from stellar irradiation.

@@

The VSI is sensitive to disk thermodynamics except
in the special case of neutrally stratified disks. It is therefore
important to understand and quantify this dependence, for example on
disk parameters, in order to evaluate the viability of the VSI in
realistic disks. Numerical simulations of the VSI are challenging,
however, because global disk simulations should be carried out while
resolving radial scales much smaller than the local disk scale height.

On the other hand, should a given disk be
unstable to the VSI, it must pass through the linear instability
first. Thus, studying the thermodynamic dependence of the linear VSI
is a useful first step. Furthermore, linear stability calculations are
essential to complement future numerical simulations.

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations to be varied smoothly between isothermal and adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of early
analysis of \cite{urpin98} and \cite{urpin03} who consider
vertically-localized disturbances. Our vertically-global approach is
necessary to study disturbances with vertical wavelengths
comparable to or larger than the disk scale height. Such disturbances were 
found by \cite{nelson13} to dominate their numerical simulations. 
We thus focus on vertically-global perturbations.

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the
  disk  midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in a stably stratified disk. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include an energy
  equation with a thermal relaxation term that restores the disk
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from
  zero can dramatically reduce VSI growth rates. % , since this generally
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  Our  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN)
  with an estimate of the physical thermal timescales in the MMSN
  based on dust opacity. We find the MMSN is unstable to the VSI at
  tens of AU with a characteristic growth times of tens of local
  orbital periods. This suggests the VSI is dynamically important in
  the outer parts of protoplanetary disks.  
\end{abstract}

@@

%recent work on vsi 
One possibility is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. The VSI is the
disk analog of the Goldreich-Schubert-Fricke instability
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple interpretation for the VSI: an
exchange of fluid parcels while conserving their angular momenta
individually, may result in a new configuration of lower energy, if
the exchange occurs over a sufficiently large vertical displacement.

@@

The VSI has been studied analytically in local approximation
\citep{urpin98,urpin03}. Numerical simulations of the  
VSI have been performed by \cite{arlt04}, and more recently by 
\cite{nelson13} and \cite{stoll14}. In these simulations the authors indeed observe the
growth of vertically-global but radially-localized perturbations when
the equilibrium rotation varied with height.

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions are stabilized by
buoyancy if the disk is sub-adiabatically stratified, as required for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P_\mathrm{orb}$, \cite{nelson13} find that
$t_c\lesssim 0.01P_\mathrm{orb}$ was needed to observe instability. That is, the
temperature profile of the disk essentially remains unchanged during
the disk evolution. This has been confirmed by 
\cite{stoll14}, who find the VSI is only sustained in radiation
hydrodynamic simulations with the inclusion of heating from stellar
irradiation, which resulted in an effective thermal timescale $t_c\ll
P_\mathrm{orb}$.

@@

From these studies it is clear that the VSI is generally sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually assumed for typical disk
models,  
% disk thermodynamics except
% in the special case of neutrally stratified disks. 
it is 
important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the viability of
the VSI in realistic disks. % Numerical simulations of the VSI are
% challenging, however, because global disk simulations should be
% carried out while resolving radial scales much smaller than the local
% disk scale height.    
%On the other hand, 
Should a given disk be unstable to the VSI, it must pass through the
linear instability first. Thus, studying the thermodynamic dependence
of the linear VSI is essential. % Furthermore, linear
% stability calculations are essential to complement future numerical simulations.

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations to be varied smoothly between isothermal and adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of early
analysis of \cite{urpin98} and \cite{urpin03} which accounts for thermal
relaxation, but considered vertically localized disturbances. Our
vertically global approach is necessary to study disturbances with
vertical wavelengths comparable to or larger than the disk scale
height. Such disturbances were  found by \cite{nelson13} and
\cite{stoll14} to dominate their numerical simulations.  We thus focus
on vertically global perturbations.
